### 2022-12-14 chapter1/chapter1.tex
Added TODOs about updating trimming method in main text and methods
Added templates for remaining chapters
Updated figure labels to match file names

% Abstract must be 300 words or less
\begin{abstract}
\noindent
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
\end{abstract}

@@

After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment. Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~\ref{fig:hmm}B). Overall, fewer than 10\% of alignments were trimmed of any regions. However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:hmm_trim}).

@@

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:hmm_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

@@

>>>>

\begin{abstract}
\noindent
Identifying protein sequences with common ancestry is a core task in bioinformatics and evolutionary biology. However, methods for inferring and aligning such sequences in annotated genomes have not kept pace with the increasing scale and complexity of the available data. Thus, in this work we implemented several improvements to the traditional methodology that more fully leverage the redundancy of closely related genomes and the organization of their annotations. Two highlights include the application of the more flexible \textit{k}-clique percolation algorithm for identifying clusters of orthologous proteins and the development of a novel technique for removing poorly supported regions of alignments with a phylogenetic HMM. In making the latter, we also wrote a fully documented Python package Homomorph that implements standard HMM algorithms and created a set of tutorials to promote its use by a wide audience. We applied the resulting pipeline to a set of 33 annotated \textit{Drosophila} genomes, generating 22,813 orthologous groups and 8,565 high-quality alignments.
\end{abstract}

@@

After the model was trained on manually labeled examples, it was used to assign a label to the columns in each alignment. Columns assigned to states 1A and 1B are the regions of interest for downstream analyses since the gaps generally follow the expected pattern given the phylogenetic tree. In contrast, columns assigned to states 2 and 3 largely corresponded to the long poorly supported segments and phylogenetically discordant regions discussed previously and were therefore removed from the alignments. In the example decoded alignment shown, the decoded states closely follow the expected patterns (Fig.~\ref{fig:hmm}B). Overall, fewer than 10\% of alignments were trimmed of any regions. However, when regions were trimmed, they were largely inferred as state 3, meaning the removed regions are generally long poorly supported segments that are aligned to few if any other sequences (Fig.~\ref{sfig:insertion_trim}).

% TODO: Update text for new figure

@@

All HMM algorithms were implemented with custom code which is available as the package Homomorph on PyPI. The insertion phylo-HMM was trained on 39,298 manually labeled columns in 11 alignments, and the deletion phylo-HMM was trained on 58,655 manually labeled positions in 20 sequences in 9 unique alignments (Fig.~\ref{sfig:insertion_training}). Because maximum-likelihood estimation of the model parameters yielded posterior decoding curves which toggled between hidden states too rapidly, the models were instead trained discriminatively~\cite{Krogh1999}. The difference, briefly, is maximum-likelihood estimation finds the parameters that best reproduce the observed distributions whereas discriminative training finds the parameters that minimize prediction error. Discriminatively trained models typically perform better in practice since real-world data are rarely fully described by the distribution specified by the model.

@@

% TODO: Update training numbers, cutoffs, and add additional state 3 check
